Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2016-02-02 cordillera-cycle.tex
Add yet more corrections from Irina...

      Basal topography is bilinearly interpolated from the ETOPO1 combined
      topography and bathymetry dataset with a~resolution of 1\,arc-min
      \citep{Amante.Eakins.2009} to the model grids. Sea level is lowered as
      a~function of time based on the Spectral Mapping Project
      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale. Ice deformation
      follows temperature and water-content dependent creep
      (Sect.~\ref{sec:icedyn}). Default parameter values are given in
      Table~\ref{tab:params}. Basal sliding follows
      a~pseudo-plastic law where the yield stress accounts for till
      deconsolidation under high water pressure (Sect.~\ref{sec:sliding}).
      Ice shelf calving is simulated using an ice thickness threshold and
      principle components of the strain rate tensor
      (Sect.~\ref{sec:calving}). Surface mass balance is
      computed using a~positive degree-day (PDD) model
      (Sect.~\ref{sec:surface}). Climate forcing is provided by a~monthly
      climatology averaged from 1979 to 2000 from the North American
      Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, perturbed
      by time-dependent offsets and lapse-rate temperature corrections
      (Sect.~\ref{sec:atm}, Table~\ref{tab:records}). A~sensitivity study
      to some of the most influential ice rheology (Sect~\ref{sec:icedyn})
      and basal sliding (Sect.~\ref{sec:sliding}) parameters was performed
      (Table~\ref{tab:sens_params}).

@@

      On the other hand, we test the model sensitivity
      (Sect.~\ref{sec:sens}) to different values for the two creep
      parameters, $A_{\mathrm{c}}$ and $A_{\mathrm{w}}$, the two activation
      energies, $Q_{\mathrm{c}}$ and $Q_{\mathrm{w}}$, and the SIA
      enhancement factor, $E_{\text{SIA}}$, as follow.
%
\begin{itemize}
  \item Our \emph{default} configuration used in the control run of the
    sensitivity study and all other simulations in the paper includes
    rheological parameters,
    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
    $Q_{\mathrm{w}}$, derived from \citet{Paterson.Budd.1982} and given in
    \citet[Eqn.~5]{Bueler.Brown.2009}, and $E_{\text{SIA}}=1$.
  \item Our \emph{hard ice} configuration includes rheological parameters,
    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
    $Q_{\mathrm{w}}$, derived from \citet[p.~72 and
    76]{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=1$, which correspond to
    a~stiffer rheology than that used in the control run.
  \item Our \emph{soft ice} configuration includes rheological parameters
    from \citet{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=5$, the
    recommended value for ice age polar ice
    \citep[p.~77]{Cuffey.Paterson.2010}.
\end{itemize}
%
      An additional simulation using the ice rheology from
      \citet{Cuffey.Paterson.2010} and $E_{\text{SIA}}=2$, the recommended
      value for Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010} was
      performed, but its results were very similar to that of our default
      run and thus not presented here.

@@

      In all our simulations, we set constant the pseudo-plastic sliding
      exponent, $q=0.25$, and the threshold velocity,
      $v_{\text{th}}=100$\,\unit{m\,a^{-1}}, according to values used by
      \citet{Aschwanden.etal.2013}, the till cohesion, $c_0=0$, whose
      measured values are consistently negligible
      \citep[p.~268]{Tulaczyk.etal.2000, Cuffey.Paterson.2010}, the till
      reference void ratio, $e_0=0.69$, and the till compressibility
      coefficient, $C_{\mathrm{c}}=0.12$, according to the only measurements
      available to our knowledge, published by \citep{Tulaczyk.etal.2000}.
      These fixed parameter values are summarized in Table~\ref{tab:params}.

      We also use a~constant spatial distribution of the till friction
      angle, $\phi$, whose values vary from 15 to 45\unit{\degree} as
      a~piecewise-linear function of modern bed elevation, with the lowest
      value occuring below modern sea level (0\,\unit{m} above sea level,
      \unit{m\,a.s.l.}) and the highest value occuring above the generalised
      elevation of the highest shorelines
      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}. This range of
      values span over the range of measured values for glacial till of 18
      to 40\unit{\degree} \citep[p.~268]{Cuffey.Paterson.2010}. It accounts
      for frictional basal conditions associated with discontinuous till
      cover at high elevations, and for a~weakening of till associated with
      the presence of marine sediments (cf. \citealp{Martin.etal.2011};
      \citealp[Supplement]{Aschwanden.etal.2013};
      \citealp{PISM-authors.2015}).

      An additional simulation with a~constant till friction angle,
      $\phi=30$\unit{\degree}, corresponding to the average value in
      \citet[p.~268]{Cuffey.Paterson.2010}, was actually performed, but the
      induced variability was small and thus not presented here.

@@

      Finally, the bedrock topography responds to ice load following
      a~bedrock deformation model that includes local isostasy, elastic
      lithosphere flexure and viscous astenosphere deformation in an
      infinite half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
      A~relatively low viscosity value of $\nu_{\mathrm{m}} =
      1\times10^{19}$\,\unit{Pa\,s} is used for the astenosphere
      (Table~\ref{tab:params}) in accordance with the results from regional
      glacial isostatic adjustment modelling at the northern Cascadia
      subduction zone \citep{James.etal.2009}.

@@

      Climate forcing driving ice sheet simulations consists of
      a~present-day monthly climatology, $\{T_{\mathrm{m}0},
      P_{\mathrm{m}0}\}$, where temperatures are modified by offset time
      series, ${\Delta}T_{\text{TS}}$, and lapse-rate corrections,
      ${\Delta}T_{\text{LR}}$:
\begin{align}
&T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) + {\Delta}T_{\text{TS}}(t)
                    + {\Delta}T_{\text{LR}}(t, x, y) \,, \\
&    P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \,.
\end{align}
      The present-day monthly climatology was bilinearly interpolated from
      near-surface air
      temperature and precipitation rate fields from the NARR, averaged from
      1979 to 2000. Modern climate of the North American Cordillera is
      characterised by strong geographic variations in temperature
      seasonality, timing of the maximum annual precipitation, and daily
      temperature variability (Fig.~\ref{fig:atm}). Although the ability
      of NARR to reproduce these steep climatic gradients is limited by its
      spatial resolution of 32\,\unit{km} \citep{Jarosch.etal.2012}, it was
      tested against observational data in a~previous sensitivity study and
      identified as yielding modelled Cordilleran ice sheet LGM extent in
      closer agreement to geological reconstructions than other atmospheric
      reanalyses \citep{Seguinot.etal.2014}.

      Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived
      from palaeo-temperature proxy records from the Greenland Ice Core
      Project \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice
      Core Project \citep[NGRIP,][]{Andersen.etal.2004}, the European
      Project for Ice Coring in Antarctica \citep[EPICA,][]
      {Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and
      Ocean Drilling Program (ODP) sites 1012 and 1020, both located off the
      coast of California \citep{Herbert.etal.2001}. Palaeo-temperature
      anomalies from the GRIP and NGRIP records were calculated from oxygen
      isotope (\chem{\delta^{18}O}) measurements using a~quadratic equation
      \citep{Johnsen.etal.1995},
\begin{align}
{\Delta}T_{\text{TS}}(t) ={~}&-11.88 [\chem{\delta^{18}O}(t)
                                -\chem{\delta^{18}O}(0)] \nonumber \\
                        &-0.1925[\chem{\delta^{18}O}(t)^2
                                 -\chem{\delta^{18}O}(0)^2] \,,
\end{align}
      while temperature reconstructions from Antarctic and oceanic cores
      were provided as such. For each proxy record used and each of the
      configurations used in the sensitivity tests, palaeo-temperature
      anomalies were scaled linearly (Tables~\ref{tab:records}
      and~\ref{tab:sens_params}) in order to simulate comparable ice extents
      at the LGM (Table~\ref{tab:extrema}) and realistic outlines
      (Fig.~\ref{fig:lr_maps}).

>>>>

      Basal topography is bilinearly interpolated from the ETOPO1 combined
      topography and bathymetry dataset with a~resolution of 1\,arc-min
      \citep{Amante.Eakins.2009} to the model grids. It responds to ice load
      following a~bedrock deformation model that includes local isostasy,
      elastic lithosphere flexure and viscous astenosphere deformation in an
      infinite half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
      A~relatively low viscosity value of $\nu_{\mathrm{m}} =
      1\times10^{19}$\,\unit{Pa\,s} is used for the astenosphere
      (Table~\ref{tab:params}) in accordance with the results from regional
      glacial isostatic adjustment modelling at the northern Cascadia
      subduction zone \citep{James.etal.2009}. Sea level is lowered as
      a~function of time based on the Spectral Mapping Project
      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale.

      Ice deformation
      follows temperature and water-content dependent creep
      (Sect.~\ref{sec:icedyn}). Default parameter values are given in
      Table~\ref{tab:params}. Basal sliding follows
      a~pseudo-plastic law where the yield stress accounts for till
      deconsolidation under high water pressure (Sect.~\ref{sec:sliding}).
      Ice shelf calving is simulated using an ice thickness threshold and
      principle components of the strain rate tensor
      (Sect.~\ref{sec:calving}). Surface mass balance is
      computed using a~positive degree-day (PDD) model
      (Sect.~\ref{sec:surface}). Climate forcing is provided by a~monthly
      climatology averaged from 1979 to 2000 from the North American
      Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, perturbed
      by time-dependent offsets and lapse-rate temperature corrections
      (Sect.~\ref{sec:atm}, Table~\ref{tab:records}). A~sensitivity study
      to some of the most influential ice rheology (Sect~\ref{sec:icedyn})
      and basal sliding (Sect.~\ref{sec:sliding}) parameters is performed
      (Table~\ref{tab:sens_params}).

@@

      On the other hand, we test the model sensitivity
      (Sect.~\ref{sec:sens}) to different values for the two creep
      parameters, $A_{\mathrm{c}}$ and $A_{\mathrm{w}}$, the two activation
      energies, $Q_{\mathrm{c}}$ and $Q_{\mathrm{w}}$, and the SIA
      enhancement factor, $E_{\text{SIA}}$, as follows.
%
\begin{itemize}
  \item Our \emph{default} configuration used in the control run of the
    sensitivity study and all other simulations in the paper includes
    rheological parameters,
    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
    $Q_{\mathrm{w}}$, derived from \citet{Paterson.Budd.1982} and given in
    \citet[Eqn.~5]{Bueler.Brown.2009}, and $E_{\text{SIA}}=1$.
  \item Our \emph{hard ice} configuration includes rheological parameters,
    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
    $Q_{\mathrm{w}}$, derived from \citet[p.~72 and
    76]{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=1$, which correspond to
    a~stiffer rheology than that used in the control run.
  \item Our \emph{soft ice} configuration includes rheological parameters
    from \citet{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=5$, the
    recommended value for ice age polar ice
    \citep[p.~77]{Cuffey.Paterson.2010}.
\end{itemize}
%
      An additional simulation using the ice rheology from
      \citet{Cuffey.Paterson.2010} and $E_{\text{SIA}}=2$, the recommended
      value for Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010} has
      been performed, but is not presented here, since its results were very
      similar to that of our default run.

@@

      In all our simulations, we set constant the pseudo-plastic sliding
      exponent, $q=0.25$, and the threshold velocity,
      $v_{\text{th}}=100$\,\unit{m\,a^{-1}}, according to values used by
      \citet{Aschwanden.etal.2013}, the till cohesion, $c_0=0$, whose
      measured values are consistently negligible
      \citep[p.~268]{Tulaczyk.etal.2000, Cuffey.Paterson.2010}, the till
      reference void ratio, $e_0=0.69$, and the till compressibility
      coefficient, $C_{\mathrm{c}}=0.12$, according to the only measurements
      available to our knowledge, published by \citet{Tulaczyk.etal.2000}.
      These fixed parameter values are summarized in Table~\ref{tab:params}.

      We also use a~constant spatial distribution of the till friction
      angle, $\phi$, whose values vary from 15 to 45\unit{\degree} as
      a~piecewise-linear function of modern bed elevation, with the lowest
      value occuring below the modern sea level (0\,\unit{m} above sea level,
      \unit{m\,a.s.l.}) and the highest value occuring above the generalised
      elevation of the highest shorelines
      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}. This range of
      values span over the range of measured values for glacial till of 18
      to 40\unit{\degree} \citep[p.~268]{Cuffey.Paterson.2010}. It accounts
      for frictional basal conditions associated with discontinuous till
      cover at high elevations, and for a~weakening of till associated with
      the presence of marine sediments (cf. \citealp{Martin.etal.2011};
      \citealp[Supplement]{Aschwanden.etal.2013};
      \citealp{PISM-authors.2015}).

      An additional simulation with a~constant till friction angle,
      $\phi=30$\unit{\degree}, corresponding to the average value in
      \citet[p.~268]{Cuffey.Paterson.2010}, has been performed, but is not
      presented here, since the induced variability was small.

@@

@@

      Climate forcing driving ice sheet simulations consists of
      a~present-day monthly climatology, $\{T_{\mathrm{m}0},
      P_{\mathrm{m}0}\}$, where temperatures are modified by offset time
      series, ${\Delta}T_{\text{TS}}$, and lapse-rate corrections,
      ${\Delta}T_{\text{LR}}$:
\begin{align}
&T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) + {\Delta}T_{\text{TS}}(t)
                    + {\Delta}T_{\text{LR}}(t, x, y) \,, \\
&    P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \,.
\end{align}
      The present-day monthly climatology was bilinearly interpolated from
      near-surface air
      temperature and precipitation rate fields from the NARR, averaged from
      1979 to 2000. Modern climate of the North American Cordillera is
      characterised by strong geographic variations in temperature
      seasonality, timing of the maximum annual precipitation, and daily
      temperature variability (Fig.~\ref{fig:atm}). Although the ability
      of the NARR to reproduce the steep climatic gradients is limited by its
      spatial resolution of 32\,\unit{km} \citep{Jarosch.etal.2012}, it has
      been tested against observational data in our previous sensitivity
      study and identified as yielding a closer fit between the modelled LGM
      extent of the Cordilleran ice sheet and the geological evidence than
      other atmospheric reanalyses \citep{Seguinot.etal.2014}.

      Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived
      from palaeo-temperature proxy records from the Greenland Ice Core
      Project \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice
      Core Project \citep[NGRIP,][]{Andersen.etal.2004}, the European
      Project for Ice Coring in Antarctica \citep[EPICA,][]
      {Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and
      Ocean Drilling Program (ODP) sites 1012 and 1020, both located off the
      coast of California \citep{Herbert.etal.2001}. Palaeo-temperature
      anomalies from the GRIP and NGRIP records were calculated from oxygen
      isotope (\chem{\delta^{18}O}) measurements using a~quadratic equation
      \citep{Johnsen.etal.1995},
\begin{align}
{\Delta}T_{\text{TS}}(t) ={~}&-11.88 [\chem{\delta^{18}O}(t)
                                -\chem{\delta^{18}O}(0)] \nonumber \\
                        &-0.1925[\chem{\delta^{18}O}(t)^2
                                 -\chem{\delta^{18}O}(0)^2] \,,
\end{align}
      while temperature reconstructions from Antarctic and oceanic cores
      were provided as such. For each proxy record used and each of the
      parameter setup used in the sensitivity tests, palaeo-temperature
      anomalies were scaled linearly (Tables~\ref{tab:records}
      and~\ref{tab:sens_params}) in order to simulate comparable ice extents
      at the LGM (Table~\ref{tab:extrema}) and realistic outlines
      (Fig.~\ref{fig:lr_maps}).
